---
title: "Simulating host parasite dynamics"
author: "Max R Brown"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: github_document

---

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Simulating host parasite dynamics

I am completely new to modelling host-parasite dynamics. I just thought it would be interesting to do, as I have been investigating this with emprical data. The object here was to simulate an interaction between a host and a parasite, to look at dynamics primarily from the point of view of the parasite. The object was also to sort of 'tailor' this for plants. Plant parasites are interesting because they cannot move; perhaps that makes the dynamics easier to code! I am frightfully bad at this, and I am really interested in taking this somewhere. I will annotate the code as fully as I can and hopefully someone, somewhere can help make these models (1) work properly, (2) under a variety of scenarios and (3) make them more efficient! They are pretty slow.


## Installation

You can install this package by using the following commands:

```{r}
# install.packages(devtools)
# devtools::install_github("Euphrasiologist/host_parasite_simulations")
library(HoPaSim)
```

## Rationale

These functions are designed to model a simple host parasite interaction with a number of assumptions. A host grid is generated, which is fixed for all future generations (i.e. they are perennial hosts). A parasite grid is then overlaid, which contains ephemeral annual plant parasites which persist only for a single generation. Depending on the benefit a host gifts the parasite, this will determine the number of progeny seeded in the next generation. Explicitly, the assumptions are:

- that hosts are fixed in position and perennial, and are unaffected by parasitism. This is probably not totally realistic, because in reality, plant parasites have the capability of being really quite virulent.
- no effect on parasite of nearby hosts. There is no additional weighted matrix which accounts for the spatial effects of parasitism. In reality, we know that plant parasites can attach to multiple hosts at the same time. Maybe for a future implementation.
- parasites seed randomly over the field. Seed disperal in parasitic plants vary. Some are dispersed by ants, others mechanically or by wind. 
- all parasites survive to end of generation if they land on a host, and selection is > 0. In the wild, parasites may well die due to environmental conditions or pathologies of host defence etc.
- parasites do not produce any offspring without a host, and all parasites with a host reproduce. This is related to the previous point, however, it is known for some species (especially facultative parasites) that seed can be produced even without a host plant species.

Given these (not unreasonable) assumptions we can create a model which can simulate parasite population sizes over time measured in generations. These can be for single host parasite systems, single parasite and multiple hosts, or multiple parasite species and multiple host species.

## Single host single parasite system

This is the most simple system to model. In the functions, the user can determine the size of the field to play on, the selective advantage measured in terms of offspring that survive to maturity in the next generation if you land on the host, number of host individuals and number of parasites. Number of generations can also be specified.

The output is extremely verbose, as each iteration of the spatial matrix structure is saved [maybe change this?]. The interesting part is probably the first slot in the output, accessed as shown below, which saves population sizes over time

```{r}
library(data.table); library(ggplot2)
sim1 <- HoPaSim::onehost_oneparasite(field.size = 25^2, s = 2, host.number = 400, parasite.number = 20, gens = 100)

sim1.2 <- setDT(sim1[[1]])

head(sim1.2)
```

A plot can then be generated using the data.

```{r}
ggplot(sim1.2, aes(x = generation, y = P.size))+geom_line() + theme_bw() + xlab(label = "Generations") + ylab(label = "Population Size")

```

For these parameters under this model, the parasite sweeps rapidly to parasitise every possible host in about 20 generations. It is capped at the upper limit of 400 as this is the number of host plants I seeded the field with.

This example can be easily replicated many times using the function `onehost_oneparasite_dynamics()`, with an extra parameter argument, reps, which can be used to specify the number of replicate simulations to use.

```{r}
# use same parameters as above
sim2 <- suppressWarnings(HoPaSim::onehost_oneparasite_dynamics(field.size = 25^2, s = 2, host.number = 400, parasite.number = 20, gens = 100, reps = 30))
head(sim2)

ggplot(sim2, aes(x = generation, y = P.size))+geom_line(aes(group = rep)) + theme_bw() + xlab(label = "Generations") + ylab(label = "Population Size")
```

From this, it is easy to see that there is certainly variability associated with this simulation, but they are all roughly doing a similar thing. The driving engine for the simulations is the R function `sample`, and this is causing the variability.